Work Fragmentation in Developer Interaction Data

### 2016-02-23 paper_2.tex
Minor changes

While these studies already show the effect of work fragmentation can be hurtful, they are mostly observational studies over a limited time, and do not address specifically the sub-population of programmers. According to Parnin and Rugaber \cite{PR12}, programmers must perform activities that challenge the memory, like keeping track of refactoring changes, navigate and debug unfamiliar code, learn new APIs, and understand new concepts; when these tasks are interrupted, it can be costly to recover the lost state. Some of these tasks require to invest in program comprehension activities, which can account from 54 to 95\% of the time spent on every working session \cite{MMLK14}. Program comprehension is especially hard, since programmers need to build a mental model of an existing program; such mental models are often very complex, involving dozens or hundreds of program elements, which is much more than can comfortably fit in memory. Such an activity is especially difficult to interrupt: Parnin and Rugaber describe the various information needs that programmer have in the situations where memory fails \cite{PR12}.

@@

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). A MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon. We built the MTS with edit and navigation variables; the time unit is the minute, and every observation is the sum of all the events that occurred in this minute. We selected the minute as unit of time because it seemed to be an appropriate and minimal representation of the user interaction in a programming task---obviously a subjective decision.

@@

In this section we analyze whether or not the interruption duration is a factor in the relationship between interruptions and developer productivity. Under the hypothesis that prolonged interruptions require more time to recover, we have built three groups of sessions based on the proportion of long interruptions. As we described previously we consider prolonged interruption are those whose duration is $\geq$12 minutes, according to previous work \cite{GM04, KaptelininN07}. Prolonged interruptions could be due to an extended interruption, such as a coffee break or a meeting.

The proportion of long interruptions is calculated as the sum of interruptions longer than 11 minutes, divided by the total of interruptions in that session. Then, we obtain the quantiles of the distribution of this proportion and use them to create 3 groups described in the following:

@@

Following the conclusions on RQ1, the number of interruptions may affect the results if we ignore that variable. So, we split every group in accordance to the number of interruptions; this time, instead of taking the quantiles of the distribution we took the terciles. Since the results in RQ1 show that as the number of interruptions increases the difference between the groups minimizes, we believe that having four more distinctive groups instead of five is also valid. Thus, for the three metrics, we created the groups low, medium and high, and every group was split into four additional groups according to the terciles of the distribution of the number on interruptions.

One of those groups contains sessions with 1 to 3 interruptions, and we decided to split it into two groups: a group of sessions with one interruption and a second group of sessions with 2 to 3 interruptions. The intention is to see if there is a clear effect between sessions with only one interruption, when this interruption is either short or prolonged. So, we created five groups in total: the \textit{none} group of sessions without interruptions, a group of sessions with only 1 interruption, another group with 2 to 3 interruptions, a fourth group with 4 to 6 interruptions, and lastly a group of sessions with more than 6 interruptions. The value of the terciles turned out to be the same for Mylyn and UDC (30\%: 3 and 60\%: 6) so we kept the same groups in both analyses.

>>>>

While these studies already show the effect of work fragmentation can be hurtful, they are mostly observational studies over a limited time, and do not address specifically the sub-population of programmers. According to Parnin and Rugaber \cite{PR12}, programmers must perform activities that challenge the memory, like keeping track of refactoring changes, navigate and debug unfamiliar code, learn new APIs, and understand new concepts; when these tasks are interrupted, it can be costly to recover the lost state. Some of these tasks require to invest in program comprehension activities, which can account from 54 to 95\% of the time spent on every working session \cite{MMLK14}.

Program comprehension is especially hard, since programmers need to build a mental model of an existing program; such mental models are often very complex, involving dozens or hundreds of program elements, which is much more than can comfortably fit in memory. Such an activity is especially difficult to interrupt: Parnin and Rugaber describe the various information needs that programmer have in the situations where memory fails \cite{PR12}. Something similar happens in the context of businessman and accountants when building and debugging spreadsheets, as they must keep the context of the information represented in terms of real world problems and therefore they must build complex mental models \cite{KS13}.

@@

\subsection{Transformation}
Our goal is to build compact and representative models from each session. In this sense, we used aggregation of events to generate a multivariate time series (MTS). A MTS is a sequence of multivariate observations taken at continuous time intervals coming from a same phenomenon and it is a way of transforming a spatial-temporal model to a simple time serie model, based on the work by Rohrdantz et al. \cite{RHD12}.

We built the MTS with edit and navigation variables; the time unit is the minute, and every observation is the sum of all the events that occurred in this minute. We selected the minute as unit of time because it seemed to be an appropriate and minimal representation of the user interaction in a programming task---obviously a subjective decision.

@@

In this section we analyze whether or not the duration of the interruptions is a factor in the relationship between interruptions and developer productivity. Under the hypothesis that prolonged interruptions require more time to recover, we built three groups of sessions based on the proportion of long interruptions. As we described previously we consider that prolonged interruption are those whose duration is $\geq$12 minutes, according to previous work \cite{GM04, KaptelininN07}. Prolonged interruptions could be due to an extended interruption, such as a coffee break or a meeting.

For every session, we calculated the proportion of prolonged interruptions as the sum of interruptions with duration equal or greater than 12 minutes, divided by the total sum of interruptions in that session. Then, we obtained the quantiles of the distribution of the proportions and use them to create 3 groups described in the following:

@@

Following the conclusions on RQ1, the number of interruptions may affect the results if we ignore that variable. So, we split every group in accordance to the number of interruptions; this time, instead of taking the quantiles of the distribution we took the terciles. Since the results in RQ1 show that as the number of interruptions increases the difference between the groups minimizes, we believe that having four more distinctive groups instead of five is also valid. Thus, for the three metrics, we created the groups low, medium and high, and then we split every group into four additional groups according to the terciles of the distribution of the number on interruptions.

One of those groups originally consisted of sessions with 1 to 3 interruptions, but we decided to split it into two groups: a group of sessions with one interruption and a second group of sessions with 2 to 3 interruptions. The intention is to see if there is a clear effect between sessions with only one interruption, when this interruption is either short or prolonged. So, we created five groups in total: the \textit{none} group of sessions without interruptions, a group of sessions with only 1 interruption, another group with 2 to 3 interruptions, a fourth group with 4 to 6 interruptions, and lastly a group of sessions with more than 6 interruptions. The value of the terciles turned out to be the same for Mylyn and UDC (30\%: 3 and 60\%: 6) so we kept the same groups in both analyses.
